---
title: "HW1"
author: "Michelle Hernandez"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
library(data.table)
library(leaflet)
library(ggplot2)
library(dplyr)
```

### Goal: to see whether daily concentrations of PM2.5 (particulate matter air pollution with aerodynamic diameter less than 2.5 `\(\mu\)`m) have decreased in California over the last 20 years (from 2002 to 2022).

## 1. Read in Data

```{r}
read_csv02 = read.csv('/Users/michellehernandez/Desktop/DataScience/HW1/HW1/ad_viz_plotval_data_2002.csv')
data_02 = data.table(read_csv02)

read_csv20 = read.csv('/Users/michellehernandez/Desktop/DataScience/HW1/HW1/ad_viz_plotval_data_2020.csv')
data_22 = data.table(read_csv20)
```

```{r}
dim(data_22)
dim(data_02)
# data from 2020 is much bigger

head(data_02)
head(data_22)

str(data_22)
str(data_02)

names(data_02)
names(data_22)


```

## 2. Combine data

```{r}
#combine data
combo = rbind(data_02, data_22)
#extract year & make year column
combo$Year <- as.integer(format(strptime(combo$Date, format = "%m/%d/%Y"), "%Y"))
#make month
combo$Month <- as.integer(format(strptime(combo$Date, format = "%m/%d/%Y"), "%m"))
```

```{r}
#rename key variables
names(combo)
names(combo)[names(combo) == "Daily.Mean.PM2.5.Concentration" ] <- "mean_pm25"
names(combo)[names(combo) == "SITE_LONGITUDE" ] <- "lon"
names(combo)[names(combo) == "SITE_LATITUDE" ] <- "lat"
names(combo)[names(combo) == "Site.ID" ] <- "site"
names(combo)[names(combo) == "DAILY_AQI_VALUE" ] <- "aqi_val"
names(combo)[names(combo) == "DAILY_OBS_COUNT" ] <- "obs_count"
names(combo)[names(combo) == "AQS_PARAMETER_CODE" ] <- "aqi_code"

```

## 3. Map

```{r}
stations <- combo %>%
  select(lat, lon, Year) %>%  
  distinct(lat, lon, .keep_all = TRUE) 

leaflet(stations) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(lat = ~lat, lng = ~lon, opacity = 1, fillOpacity = 1, radius = 400,  color = ~ifelse(stations$Year == 2002, "red", ifelse(stations$Year == 2022, "blue", "black")))
```

Looking at the map we can see that the sites for PM2.5 levels do vary by year. So not all the exact same sites are being compared, there are some new stations for 2022 and some in 2002 that were not included in 2022.

## 4. Check for missing

```{r}
max_pm <- subset(combo, mean_pm25 == 302.5)
print(max_pm)

large_pm <- subset(combo, mean_pm25 > 10.2)
print(large_pm)

large_100pm <- subset(combo, mean_pm25 > 100)
print(large_100pm)

pm0 <- subset(combo, mean_pm25 < 0)
print(pm0)


#remove negtive & NAs for PM25 values

filter_combo <- combo %>%
  filter(!is.na(mean_pm25) & mean_pm25 > 0)


```

There are some negative values which do not make sense for the PM2.5 levels. There does not appear to be a pattern to the missing values so we will remove those and continue our analysis.

```{r}
summary(filter_combo$mean_pm25)
boxplot(filter_combo$mean_pm25, main='Raw PM2.5 Distribution for all years and all stations', ylab='PM 2.5 level', xlab='all stations all years')
hist(filter_combo$mean_pm25, main='Raw PM2.5 Histogram for all years and all stations', xlab='PM2.5 level', ylab='Frequency')

```

```{r}
#Plot 1
ggplot(filter_combo, aes(x = Date, y = mean_pm25, 
      color = factor(Year), group = factor(Year))) + 
      geom_line()
```

```{r}
monthly_means <- filter_combo %>%
  group_by(Year, Month) %>%
  summarize(mean_pm25 = mean(mean_pm25, na.rm = TRUE))

# Create the plot
ggplot(monthly_means, aes(x = Month, y = mean_pm25, color = factor(Year), group = factor(Year))) + 
  geom_line() +
  scale_x_continuous(breaks = 1:12) +  
  labs(x = "Month", y = "Mean PM2.5") 
```

```{r}
# 2022 only
filtered_combo22 <- filter_combo %>% filter(Year == 2022)

ggplot(filtered_combo22, aes(x = Date, y = mean_pm25, color = factor(Year), group = factor(Year))) + 
  geom_line() +
  labs(x = "Date", y = "Mean PM2.5")

# 2002 only
filtered_combo02 <- filter_combo %>% filter(Year == 2002)

ggplot(filtered_combo02, aes(x = Date, y = mean_pm25, color = factor(Year), group = factor(Year))) + 
  geom_line() +
  labs(x = "Date", y = "Mean PM2.5")
```

## 5. Explore.

Create exploratory plots (e.g. boxplots, histograms, line plots) and summary statistics that best suit each level of data. Be sure to write up explanations of what you observe in these data. The primary question you will answer is whether daily concentrations of PM2.5 (particulate matter air pollution with aerodynamic diameter less than 2.5 `\(\mu\)`m) have decreased in California over the last 20 years (from 2002 to 2022).

state

county

site in Los Angeles

```{r}
length(unique(combo$COUNTY))
#there are 51 counties

la_combo <- subset(filter_combo, COUNTY == 'Los Angeles')
length(unique(la_combo$site))
#there are 18 sites in los angeles


```

```{r}
#State Level
ggplot(filter_combo, aes(x = STATE, y = mean_pm25, fill = factor(Year))) +
  geom_boxplot() +
  labs(x = "State", y = "PM2.5 Levels", fill = "Year") +
  ggtitle("Boxplot of PM2.5 Levels by State (2002 vs. 2022)")

hist(combo$mean_pm25[combo$Year=='2022'])



# Create the histogram plot
hist_data02 <- combo$mean_pm25[combo$Year == '2002']
hist(hist_data02,
     main = 'Histogram of Mean PM2.5 in 2002',
     ylab = 'Frequency',
     xlab = 'Mean PM2.5 level')

hist_data22 <- combo$mean_pm25[combo$Year == '2022']
hist(hist_data22,
     main = 'Histogram of Mean PM2.5 in 2022',
     ylab = 'Frequency',
     xlab = 'Mean PM2.5 level')

```

```{r}
# county level
ggplot(la_combo, aes(x = factor(Year), y = mean_pm25, fill = factor(Year))) +
  geom_boxplot() +
  labs(x = "Year", y = "PM2.5 Levels", fill = "Year") +
  ggtitle("Boxplot of PM2.5 Levels in Los Angeles County (2002 vs. 2022)")


hist_county_02 <- combo$mean_pm25[combo$Year == '2002']
hist(hist_county_02,
     main = 'Histogram of Mean PM2.5 in 2002 in LA County',
     ylab = 'Frequency',
     xlab = 'Mean PM2.5 level')

hist_county_22 <- combo$mean_pm25[combo$Year == '2022']
hist(hist_county_22,
     main = 'Histogram of Mean PM2.5 in 2022 in LA County',
     ylab = 'Frequency',
     xlab = 'Mean PM2.5 level')

```

Is this graph we can see that there is much more variation in PM2.5 level in 2002 compared to 2022 for LA County. There is also a long right tail in the histogram for 2002 which we do not see in 2022.

```{r}
la_county_data <- filter_combo %>%
  filter(COUNTY == 'Los Angeles')

la_county_data_2002 <- la_county_data %>%
  filter(Year == 2002)


la_county_data_2022 <- la_county_data %>%
  filter(Year == 2022)


summary_stats_2002 <- la_county_data_2002 %>%
  group_by(site) %>%
  summarize(
    mean_pm25_mean_2002 = mean(mean_pm25),
    mean_pm25_median_2002 = median(mean_pm25),
    mean_pm25_min_2002 = min(mean_pm25),
    mean_pm25_max_2002 = max(mean_pm25),
    mean_pm25_sd_2002 = sd(mean_pm25)
  )


summary_stats_2022 <- la_county_data_2022 %>%
  group_by(site) %>%
  summarize(
    mean_pm25_mean_2022 = mean(mean_pm25),
    mean_pm25_median_2022 = median(mean_pm25),
    mean_pm25_min_2022 = min(mean_pm25),
    mean_pm25_max_2022 = max(mean_pm25),
    mean_pm25_sd_2022 = sd(mean_pm25)
  )

summary_stats_combine <- full_join(summary_stats_2002, summary_stats_2022, by = "site")


summary_stats_combined<- summary_stats_combine %>%
  filter(!is.na(mean_pm25_mean_2002)) %>%
  filter(!is.na(mean_pm25_mean_2022))

print(summary_stats_combined)

different_sites <- summary_stats_combined %>%
  filter(is.na(mean_pm25_mean_2002) | is.na(mean_pm25_mean_2022))

print(different_sites)


#los angeles sites


# 2022
la_county_data_2002

monthly_mean_pm25_la_county_2022 <- la_county_data_2002 %>%
  group_by(Month) %>%
  summarize(mean_pm25 = mean(mean_pm25, na.rm = TRUE))

monthly_mean_pm25_la_county_2022

#2002
la_county_data_2022

monthly_mean_pm25_la_county_2002 <- la_county_data_2022 %>%
  group_by(Month) %>%
  summarize(mean_pm25 = mean(mean_pm25, na.rm = TRUE))

monthly_mean_pm25_la_county_2002




combined_data <- bind_rows(
  monthly_mean_pm25_la_county_2002 %>% mutate(Year = 2002),
  monthly_mean_pm25_la_county_2022 %>% mutate(Year = 2022)
)


ggplot(combined_data, aes(x = Month, y = mean_pm25, color = factor(Year))) +
  geom_line() +
  scale_x_continuous(breaks = 1:12) +  
  labs(x = "Month", y = "Mean PM2.5 Levels", color = "Year") +
  ggtitle("Monthly Mean PM2.5 Levels in Los Angeles County (2002 vs. 2022)")
```

```{r}
mean_pm25_by_year_month <- filter_combo %>%
  group_by(Year, Month, Year2022 = ifelse(Year == 2022, "2022", "2002")) %>%
  summarize(Mean_PM2.5 = mean(mean_pm25))


ggplot(mean_pm25_by_year_month, aes(x = Month, y = Mean_PM2.5, fill = Year2022)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_continuous(breaks = 1:12) +
  facet_wrap(~Year, nrow = 1) +
  labs(title = 'Mean PM2.5 Level by Month, for each year', x = "Month", y = "Mean PM2.5") +
  scale_fill_manual(values = c("2002" = "blue", "2022" = "red"), name = "Year") +
  theme_minimal()

```

Here we can clearly see that the mean level of PM2.5 is higher in 2002 for every month compared to 2022.

```{r}
mean_pm25_by_year <- filter_combo %>%
  group_by(Year) %>%
  summarize(Mean_PM2.5 = mean(mean_pm25))


ggplot(mean_pm25_by_year, aes(x = factor(Year), y = Mean_PM2.5, fill = factor(Year))) +
  geom_bar(stat = "identity", width = 0.7) +
   labs(title = 'Mean PM2.5 Level for Each Year', x = "Year", y = "Mean PM2.5") +
  scale_fill_manual(values = c("2002" = "blue", "2022" = "red")) 
```

The overall mean PM2.5 level is much larger in 2002 than 2022.

```{r}
# t tests

# state
t.test(filter_combo$mean_pm25[filter_combo$Year=='2002'], filter_combo$mean_pm25[filter_combo$Year=='2022'])


```

```{r}
# t test

#county
t.test(filter_combo$mean_pm25[filter_combo$Year=='2002' & filter_combo$COUNTY=='Los Angeles'], filter_combo$mean_pm25[filter_combo$Year=='2022'& filter_combo$COUNTY=='Los Angeles'])
```

It appears that PM2.5 levels have decreased from 2002 to 2022. Looking at the raw data we can see that 2022 seems to have higher PM2.5 levels, specifically larger extreme values. However if you look at the line plot and bar plots comparing mean pm2.5 values you can see that 2002 has a much higher mean for each month compared to 2022. This is apparent at the state level and count level and for the means for the whole year. A t test was also conducted at both the state and county levels. Therefore we can conclude that the mean PM2.5 levels are statistically different from 2002 to 2022 at both the state and county levels (p\<0.0001 for both). The overall conclusion that we make from the analysis is that mean PM2.5 levels decreased from 2002 to 2022.
